---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# otpr


<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/otpr)](https://cran.r-project.org/package=otpr)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/grand-total/otpr)](https://cran.r-project.org/package=otpr)
[![Build Status](https://travis-ci.org/marcusyoung/otpr.svg?branch=master)](https://travis-ci.org/marcusyoung/otpr)
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)

<!-- badges: end -->

## Overview

**otpr** is an R package that provides a simple wrapper for the [OpenTripPlanner](https://www.opentripplanner.org/) (OTP) API.
To use **otpr** you will need a running instance of OTP. The purpose of the package is
to submit a query to the relevant OTP API resource, parse the OTP response and return
useful R objects.

This package will be useful to researchers and transport planners who want to use OTP to generate trip data for accessibility analysis or to derive variables for use in transportation models.

## Installation

``` r
# Install from CRAN
install.packages("otpr")
```

### Development version

To get a bug fix, or use a feature from the development version, you can install otpr from GitHub. See [NEWS](https://github.com/marcusyoung/otpr/blob/master/NEWS.md) for changes since last release.

``` r
# install.packages("devtools")
devtools::install_github("marcusyoung/otpr")
```


## Getting started

```{r}
library(otpr)
```

### Defining an OTP connection

The first step is to call `otp_connect()`, which defines the parameters needed to connect 
to a router on a running OTP instance. The function can also confirm that the
router is reachable.

``` {r}
# For a basic instance of OTP running on localhost with standard ports and a 'default' router
# this is all that's needed
otpcon <- otp_connect()

```

## Function behaviour

The functions that query the OTP API return a list of two elements. The first
element is an errorId - with the value "OK" or the error code returned by OTP. If
errorId is "OK", the second element will contain the query response; otherwise it
will contain the OTP error message.

### Distance between two points

To get the trip distance in metres between an origin and destination on the street
and/or path network use `otp_get_distance()`. You can specify the required mode: 
CAR (default), BICYCLE or WALK are valid.

``` {r}
# Distance between Manchester city centre and Manchester airport by CAR
otp_get_distance(
  otpcon,
  fromPlace = c(53.48805,-2.24258),
  toPlace = c(53.36484,-2.27108)
)

# Now for BICYCLE
otp_get_distance(
  otpcon,
  fromPlace = c(53.48805,-2.24258),
  toPlace = c(53.36484,-2.27108),
  mode = "BICYCLE"
)

```

### Time between two points

To get the trip duration in seconds between an origin and destination use `otp_get_times()`. 
You can specify the required mode: TRANSIT (all available transit modes), BUS, RAIL, CAR, 
BICYCLE, and WALK are valid. All the public transit modes automatically allow WALK.
There is also the option to combine TRANSIT with BICYCLE.

``` {r}
# Time between Manchester city centre and Manchester airport by BUS
otp_get_times(
  otpcon,
  fromPlace = c(53.48805,-2.24258),
  toPlace = c(53.36484,-2.27108),
  mode = "BUS"
)


# By default the date and time of travel is taken as the current system date and
# time. This can be changed using the 'date' and 'time' arguments
otp_get_times(
  otpcon,
  fromPlace = c(53.48805,-2.24258),
  toPlace = c(53.36484,-2.27108),
  mode = "TRANSIT",
  date = "11-25-2018",
  time = "08:00:00"
)

```

### Breakdown of time by mode, waiting time and transfers

To get more information about the trip when using transit modes, `otp_get_times()` 
can be called with the 'detail' argument set to TRUE. The trip duration (minutes) is then further
broken down by time on transit, walking time (from/to and between stops), waiting time
(when changing transit vehicle or mode), and number of transfers (when changing transit vehicle or mode).

``` {r}
# Time between Manchester city centre and Manchester airport by TRANSIT with detail
otp_get_times(
  otpcon,
  fromPlace = c(53.48805,-2.24258),
  toPlace = c(53.36484,-2.27108),
  mode = "TRANSIT",
  date = "11-25-2018",
  time = "08:00:00",
  detail = TRUE
)

```

### Travel time isochrones

The `otp_get_isochrone()` function can be used to get one or more travel time isochrones
in GeoJSON format. These are only available for transit or walking modes (OTP limitation).
They can be generated either *from* (default) or *to* the specified location.

``` {r}
# 900, 1800 and 2700 second isochrones for travel *to* Manchester Airport by any transit mode
my_isochrone <- otp_get_isochrone(
  otpcon,
  location = c(53.36484, -2.27108),
  fromLocation = FALSE,
  cutoffs = c(900, 1800, 2700),
  mode = "TRANSIT"
)

# function returns a list of two elements
names(my_isochrone)

# now write the GeoJSON (in the "response" element) to a file so it can be opened in QGIS (for example)
write(my_isochrone$response, file = "my_isochrone.geojson")

```


## Learning more

The example function calls shown above can be extended by passing additional parameters to the OTP API. Further information is available in the documentation for each function: 

``` {r eval=FALSE}
# get help on using the otp_get_times() function
?otp_get_times
```

If you are new to OTP, then the best place to start is to work through the tutorial, [OpenTripPlanner Tutorial - creating and querying your own multi-modal route planner](https://github.com/marcusyoung/otp-tutorial). This includes
everything you need, including example data. The tutorial also has examples
of using **otpr** functions, and helps you get the most from the package, for example using it to populate
an origin-destination matrix.

For more guidance on how **otpr**, in conjunction with OTP, can be used to generate data for input into models, read [An automated framework to derive model variables from open transport data using R, PostgreSQL and OpenTripPlanner](https://eprints.soton.ac.uk/389728/).

## Getting help

* Please [report any issues or bugs](https://github.com/marcusyoung/otpr/issues).
* Get citation information for **otpr** using: `citation(package = 'otpr')`.

